### 2021-07-07 ch5langevin.tex
very close

Bulk bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
Due to these fluctuations, measured transport rates often show slow convergence through time \citep{Dhont2018,Turowski2010}, and predicted rates can deviate from measured values by several orders of magnitude \citep{Recking2012,Martin2003}.
These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{Gaeuman2017,Malmon2005}.
In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations \citep{Ancey2008} and quantifying the dependence of measurements on the observation scale \citep{Turowski2010}. 
Recent indications that sediment transport fluctuations might explain longstanding and unsolved problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume \citep{Ancey2008,Furbish2012}.
This formulation requires the instantaneous velocity distributions of sediment particles \citep{Lajeunesse2010}, although our understanding of these distributions unfortunately remains limited.
We have as of yet no consensus on the shape of the bedload velocity distribution, and although models have described some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, we have as of yet no mechanistic models that describe the full range of experimental observations \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
In this chapter, I develop a stochastic model of particle velocities which addresses this shortcoming.

Particle trajectories are a compromise between turbulent drag and particle-bed collision \citep{Wiberg1985}. We therefore anticipate that the Stokes number, which is known to characterize colliding bodies within a viscous fluid \citep{}, is an important dimensionless parameter of the bedload velocity distribution.
The Stokes number is defined as ... where ....

High-speed video experiments have measured different streamwise particle velocity distributions without providing much understanding as to why one distribution or another appears in a given set of hydraulic and sedimentary conditions.One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}.
These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) having typical Stokes numbers $St \sim 1-10$. Their flow conditions are generally sub-critical ($Fr<1$) and turbulent ($Re>5000$) but not always: flows in \citet{Lajeunesse2010} were super-critical ($Fr>1$), and flows in \citet{Charru2004} and \citet{Seizilles2014} were viscous.
A second set of studies have shown Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}. In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads having higher Stokes numbers ($St \sim 10-500$). In all Gaussian cases, flows are turbulent ($Re>5000$) and super-critical ($Fr>1$).
Two other experiments display velocity distributions that are intermediate between exponential and Gaussian and appear more like a Gamma distribution \citep{Houssais2012, Liu2019}. The \citet{Houssais2012} experiments involved a binomal distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions. They resolved the velocity distributions for larger grains only. The \citet{Liu2019} experiments used uniformly-graded sand having median diameter $1.1$mm. Flows were again turbulent and subcritical.
From this experimental record, we can summarize that the shape of the velocity distribution does not consistentely relate to whether a flow is super or sub-critical ($Fr$), whether sediment grains are natural (sand, gravel) or synthetic (beads), or whether the flow is laminar or turbulent ($Re$). However, the typical Stokes numbers of particles do seem to increase monotonically from exponential velocity experiments (where $St \sim 1$), to intermediate (Gamma-like) experiments (where $St \sim 10$), and eventually to Gaussian experiments (where $St \sim 10^2$.) Apparently, the shape of streamwise bed load velocity distributions depends on the particle size.

Existing models of streamwise bed load velocities can be divided into computational and statistical physics categories.
Computational models numerically integrate some approximate coupled dynamics for individual grains and the fluid, generally modelling particles as spheres interacting through repulsive forces, and the flow using direct simulation of the Navier-Stokes equations or some related approximation (such as large eddy simulation or the St-Venant equations). When streamwise particle velocities have been analyzed in such simulations, they show exponential tails \citep{Gonzalez2017,Furbish2013} that agree with only a subset of the experimental data.
Statistical physics models have incorporated stochastic driving and resisting terms into the Newtonian dynamics of individual grains to develop a Langevin-like description of bed load particle motions.
For the downstream velocity $u(t)$, \citet{Fan2014} wrote $\dot{u} = F -\gamma \text{sgn}{u} + \xi(t)$ where $F$ represents the steady component of the fluid forcing, the term involving $\gamma$ is a quasi-static (Coulomb-type) friction term representing momentum dissipation by particle-bed collisions, and $\xi(t)$ is a Gaussian white noise representing variability in these forces. This model provides exponential velocity distributions which agree with one subset of experiments. \citet{Ancey2014} took a similar approach that includes different forces, solving $\dot{u} = \gamma(\overline{u}-u) + \xi(t)$. Here the term involving $\gamma$ is similar to a Stokes drag, except it involves the mean sediment velocity $\overline{u}$, not the fluid velocity as for a ``real" Stokes drag. $\xi(t)$ is again a Gaussian white noise representing fluctuations, and the model provides Gaussian velocity distributions that agree with another subset of experiments. 
While these models build insight into grain-scale sediment transport mechanics and provide powerful techniques with which to approach the problem, they have not yet provided a comprehensive explanation for the range of streamwise velocity distributions resolved in experiments.

Here, motivated by the realization that experimental particle velocity distributions vary systematically with the grain size, as summarized above, we hypothesize that the shape of the streamwise velocity distribution is controlled by the momentum dissipation characteristics of particle-bed collisions.
It is well-known that the elasticity of granular collisions within viscous fluids depends on the particle size.
In addition, the velocity distributions of granular gases are known to develop increasingly heavier tails than the Gaussian (Boltzmann) form of elastic gases as the inelastiscity of particle-particle collisions is increased.
Taking inspiration from this established knowledge, we develop below a model for sediment grains in transport as they undergo inelastic particle-bed collisions. Our intentions are to test the hypothesis that particle-bed collision characteristics explain the range of experimentally-observed streamwise bed load particle velocity distributions, and to introduce more realistic forces into earlier statistical physics descriptions of individual bed load particle dynamics. 
We develop the model and explain our assumptions in section \ref{sec:model}, then we present the analytical solution and major results in \ref{sec:results}. We finally discuss the implications of these results, summarize our findings, and suggest ideas for further research in sections \ref{sec:discussion} and \ref{sec:conclusion}.

\section{Mechanistic description of particle velocities}
\label{sec:model}

Figure \ref{fig:1} indicates the configuration we have in mind. Nearly spherical and cohesionless particles of diameter $d$ and mass $m$ moving as bed load down a slope inclined at an angle $\theta$ in a steady turbulent shear flow. The flow is just strong enough to drive grains into rarefied transport of a kind typical in gravel-bed rivers: particles saltate along the bed in sequences of collisions between events of erosion and deposition; moving particles collide often with stationary particles, but rarely with other moving particles.
Particles respond to turbulent drag forces $F_D(t)$ and episodic particle-bed collision forces. In contrast to the computational physics approach, we do not aim to characterize the exact timeseries of the forces on an individual particle. Instead, we model the ensemble of possible force timeseries that particles could conceivably experience. Each possibility implies a different velocity timeseries $u(t)$ in the downstream direction.
Our objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
We include the most realistic article-bed collision and fluid forces we can while still allowing for analytical solutions.

Collision forces dissipate streamwise momentum, partly by converting it to vertical, lateral, or rotational momentum, and partly by deforming particles and generating heat \citep{Williams2021}.
The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}. Here, we introduce a restitution-like coefficient $\varepsilon$ as indicated in figure \ref{fig:1}. This ranges from $\varepsilon=0$ for completely inelastic collisions to $\varepsilon=1$ for completely elastic collisions.
If the streamwise velocity just prior to a collision is $u$, just after the collision it becomes $\varepsilon u$. 
\begin{figure}
	\centerline{\includegraphics{./figures/ch5/Fig1Concept.png}}
	\caption{Definition sketch of rarefied sediment transport with turbulent fluid drag and particle-bed collision forces. During saltation, pre-collisional streamwise velocities $u$ are transformed to postcollisional velocities $\ve u < u$.}
	\label{fig:fig1}
\end{figure}
Since this quantity combines effects of particle shape and collision geometry and should vary from one collision to the next, we consider that the fraction of streamwise momentum dissipated per collision $\varepsilon$ lies on a statistical distribution $\rho(\varepsilon)$.
Similar ideas are available in the granular physics literature \citep{Serero2015}.
Further assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep{Gordon1972}, we write the collision force in the downstream direction as
\be F_C(u,t) = - m u \sum_{k=1}^{N_\nu(t)}(1-\varepsilon_k)\delta(t-\tau_k). \label{eq:col} \ee
Here, $N_\nu(t)$ is the number of collisions in time $t$, the $\tau_k$ ($k=1,2,\dots$) are times at which collisions occur, and the $\varepsilon_k$ are elasticity coefficients characterizing the amount by which each collision slows the particle down.
This collision force is a sequence of random impulses which are proportional to the pre-collisional streamwise momentum. This collision model should be adequate when the contact times between moving and resting particles are small compared to the times between collisions. These conditions are always satisfied for the idealized saltation-type motion depicted in figure \ref{eq:1}.

Fluid forces on a coarse particle in a viscous flow depend on the Reynolds number $\Re_p = d V/\nu$ defined by the particle size $d$, slip velocity $V$ between particle and fluid, and kinematic viscosity $\nu$.
These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Re_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
At realistic $\Re_p$ analytical results are limited, so it is standard practice to turn instead to empirical corrections on the small $\Re_p$ formulas \citep{Schmeeckle2007,Clift1978}.
A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Re_p$ can be written $F_D = \frac{\pi}{8}
\rho_f d^2 C_D(\Re_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Re_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
In the present model we set $C_D = \frac{24}{\Re_p}( 1 + 0.194 \Re_p^{0.631})$ \citep{Clift1978,Gonzalez2017} and we do not involve acceleration and history terms for simplicity, although we acknowledge their potential importance for coarse sediment transport \citep{Michaelides1997,Armenio2001}.

Drag forces have been argued to fluctuate rapidly compared to the inertial response times of coarse sediment grains \citep{Fan2014}.
The magnitude of drag fluctuations has been observed to follow a Gaussian distribution \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
Using these ideas, we make two key simplifications of the drag force above.
First, we split the drag $F_D$ into quasi-steady and fluctuating components \citep{Michaelides1997}, and second, we represent drag fluctuations as a Gaussian white noise characterized by a particle diffusivity $D$ \citep{Fan2014,Ancey2014}. 
Defining $\bar{V}$ as a representative slip velocity which we specify more carefully later,  $\bar{C}_D$ as the empirical drag coefficient evaluated at this slip velocity, and $\xi(t)$ as a Gaussian white noise of mean $0$ and variance $1$ \citep{Gardiner1983}, we express the fluid forces as
\be F_D(t) = \frac{\pi}{8}
\rho_f d^2 \bar{C}_D \bar{V}^2 + \sqrt{2 D } \eta(t). \ee
In this drag force \ref{eq:drag} and the collision force \ref{eq:col}, the turbulent fluctuations $\xi(t)$, collision times $\tau_k$, and dissipation coefficients $\varepsilon_k$, can take any values consistent with their distribution and correlation functions. This set of possiblities defines a statistical ensemble.

\subsection{Langevin equation for collisional transport}

With the above forces, the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics becomes
\be m \dot{u}(t) = \Gamma + \sqrt{2D}\eta(t) - m u(t) \xi_{\nu, \ve}(t). \label{eq:langevin} \ee
This equation replaces the steady friction terms of earlier models () with an episodic term that provides a more realistic representation of particle-bed collisions during saltation.
It represents a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
Collisions introduce ``jumps" in velocity while turbulent generates "diffusion", and the collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Daly2010,Mau2014,Dubkov2019}.
One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
Here, the Ito interpretation (lower endpoint integration rule) is the physical choice since the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional.
Given this integration rule, the remaining issues are to obtain the integro-differential equation characterizing the ensemble of velocities defined by \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.

@@

The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix \ref{sec:appendixA}, providing
\be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + I_c(u,t). \label{eq:master} \ee
In this equation, we introduced the scaled parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$. The term
\be \mathcal{I}_c(u,t) = - P(u,t) + \int_0^1 \frac{d\ve}{\ve}P\big(\frac{u}{\ve},t \big) \rho(\ve) \label{eq:colint} \ee
is a ``collision integral" term representing particle-bed collisions.
Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bed load models \citep{Fan2014,Ancey2014}. 
Such equations combining are known as Chapman-Komogorov equations \citep{Gardiner1983}. 
Nonlocality is introduced by the collision integral \ref{eq:colint} which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
This term is analogous to the collision integral within the Boltzmann equation of kinetic theory \citep{Duderstadt1979, Brilliantov2004}. Physically, it corresponds to binary collisions between particles having different masses and random resitution coefficients \cite{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) goes to infinity. Mathematically, the collision integral term represents the probability distribution of the product between $\varepsilon$ and $u$ \citep[c.f.][]{Feller1968}.

Owing to its nonlocality, equation \ref{eq:master} does not admit analytical solutions as is, so I introduce one further approximation.
Assuming the distribution of dissipation coefficients $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$ allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
Expanding all terms in the integrand except $\rho(\varepsilon)$ provides
\be \mathcal{I}_c(u,t) = -P(u,t) + \frac{1}{\varepsilon'}P\big(\frac{u}{\ve'},t \big) + \sum_{k=1}^\infty \frac{\alpha_k }{k!}(\ve - \ve')^k \Big[\frac{1}{\ve}P\big(\frac{u}{\ve}\big)\Big]^{(k)}\Big|_{\ve=\ve'},\label{eq:expansion}\ee
where the $\alpha_k = \int_0^1 d\ve \rho(\ve) (\ve-\ve')^k $ are the central moments of $\ve$ around the mode elasticity $\ve'$ and the superscript $(k)$ denotes the $k$th derivative.
In what follows, I drop all but the first two terms in this expansion to obtain the leading order contribution of particle-bed collisions to the velocity distribution. Higher orders could always be included later by perturbation theory \citep{Morse1953}. I solve the resulting approximate equation in steady-state, when $\partial P(u,t)/\partial t = 0$. Equation \ref{eq:master} indicates that this solution will be a good approximation to the time-dependent problem when particle motions generally survive multiple collisions ($t\gg1/\nu$).

\section{Results}
\label{sec:results}

>>>>

Bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
Owing to these fluctuations, measured transport rates often show slow convergence through time \citep{Dhont2018,Turowski2010}, and predicted rates can deviate from measured values by several orders of magnitude \citep{Recking2012,Martin2003}.
These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{Gaeuman2017,Malmon2005}.
In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations and quantifying the dependence of measurements on the observation scale (secs. \ref{sec:birthdeath}-\ref{sec:renewal}).
Recent indications that sediment transport fluctuations might explain longstanding open problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019,Abramian2020} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume (sec. \ref{sec:birthdeath}).
This formulation requires the instantaneous velocity distributions of sediment particles \citep{Lajeunesse2010}, although understanding of these distributions remains limited.
There is as of yet no consensus on the shape of the bedload velocity distribution, and although the models described in section \ref{sec:langevin} of the introduction have described some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, no models have been developed that describe the full range of experimental observations \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
In this chapter, I develop a stochastic model of particle velocities which addresses this shortcoming.

High-speed video experiments have measured different streamwise particle velocity distributions without indicating why one distribution or another appears in a given set of hydraulic and sedimentary conditions.
One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}.
These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) in turbulent and subcritical flows, but not always; the experiments of \citet{Lajeunesse2010} were supercritical, while the experiments of \citet{Charru2004} and \citet{Seizilles2014} were viscous.
A second set of studies have shown Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}. In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads, and flows are generally turbulent and super-critical.
Other experiments display velocity distributions that are intermediate between exponential and Gaussian extremes that appear visually more like a Gamma distribution \citep{Houssais2012, Liu2019}.
The \citet{Houssais2012} experiments involved a binomal distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions, while the \citet{Liu2019} experiments used sand having median diameter $1.1$mm. Flows were again turbulent and subcritical.
From this experimental record, the velocity distribution shape apparently does not consistently relate to super or sub-critical flow conditions, whether sediment grains are natural (sand, gravel) or synthetic (glass beads), or whether the flow is laminar or turbulent.
However there is a loose trend within the particle size. Smaller particles tend to show more exponential-like distributions \citep[e.g.][]{Fathel2015}, while larger ones show Gaussian distributions \citep[e.g.][]{Heyman2016}. This summary suggests that the shape of streamwise bed load velocity distribution could be controlled by the particle size.

Existing models of streamwise bed load velocities can be divided into computational and stochastic physics categories.
Computational models numerically integrate some approximate coupled dynamics for individual grains and the fluid, generally modelling particles as spheres interacting through repulsive forces, and the flow using direct simulation of the Navier-Stokes equations or some related approximation (such as large eddy simulation or the St-Venant equations).
When streamwise particle velocities have been analyzed in such simulations, they show exponential tails \citep{Gonzalez2017,Furbish2013} that agree with one subset of the experimental data.

Several stochastic models have incorporated fluctuating driving and resisting terms into the Newtonian dynamics of individual grains to develop Langevin-like descriptions of bed load particle motions (sec. \ref{sec:langevin}). \citet{Fan2014} represented turbulent drag as Gaussian white noise and included a static Coulomb friction term to model particle-bed collisions, while \citet{Ancey2014} applied an Ornstein-Uhlenbeck process which combines the fluid and driving forces were lumped into a velocity-dependent force, while Gaussian white noise represents fluctuations.
Each of these models derives one end-member distribution from among the range of distributions reported in experiments.
A physical description for the full range of experimentally-observed bedload velocity distributions remains an elusive target.

In this chapter I explore the possibility that the shape of the particle velocity distribution is controlled by momentum dissipation during particle-bed collisions.
In the theory of granular gases, dissipative collisions are known to cause departures toward an exponential velocity distribution from the ideal Gaussian form expected from perfectly elastic collisions \citep{Brilliantov2004}.
Gas theory formulates particle-particle collisions as a nonlocal integral within the master equation for the probability distribution of particle velocity called the Boltzmann equation \citep{Chapman1970,Landau1980}.
Collisions in these models involve episodic kicks to the particle momenta at random times, parameterized by billiard ball-like models of the underlying rigid body dynamics \citep[e.g.][]{Brach1989}.
In contrast, current stochastic models of bedload transport appear highly simplified, treating collisions instead with static Coulomb or velocity-dependent drag.

Taking inspiration from granular gas theory, I develop in this chapter an improved model for sediment grains in transport, including episodic particle-bed collisions.
The driving motivation is to test the hypothesis that particle-bed collision characteristics can explain the range of experimentally-observed streamwise bed load particle velocity distributions.
A secondary one is to introduce more realistic forces into earlier stochastic descriptions of individual bed load particle dynamics. 
I develop the model and explain its key assumptions in section \ref{sec:langlangmodel}. Then I present the analytical solution of the model and other major results in \ref{sec:langresults}. Finally I discuss the implications of these results, summarize key findings, and suggest ideas for further research in sections \ref{sec:langdiscussion} and \ref{sec:langconclusion}.

\section{Mechanistic description of particle velocities}
\label{sec:langmodel}

Figure \ref{fig:1} indicates the configuration of the model in this chapter.
Nearly spherical cohesionless particles of diameter $d$ and mass $m$ move as bed load down a slope inclined at an angle $\theta$ in a steady turbulent shear flow.
The flow is just strong enough to drive grains into the rarefied transport typical in gravel-bed rivers: particles saltate downstream through a sequence of particle-bed collisions. 
Moving particles collide often with stationary particles, but rarely with other moving particles.

Particles respond to turbulent drag forces $F_D(t)$ and episodic particle-bed collision forces. In contrast to the computational physics approach, I do not aim to characterize the exact timeseries of the forces on an individual particle. Instead, I model the ensemble of possible force timeseries that particles could conceivably experience. Each possibility implies a different possible velocity timeseries $u(t)$ in the downstream direction.
The objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
I include the most realistic article-bed collision and fluid forces possible while still allowing for analytical solutions.

\subsection{Episodic collisions}
Collision forces dissipate streamwise momentum, partly by converting it to vertical, lateral, or rotational momentum, and partly by deforming particles and generating heat \citep{Schmeeckle2014,Williams2021}.
The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}. Here, we introduce a simple elasticity parameter $\varepsilon$ as a first stab at the problem, as indicated in figure \ref{fig:1}. This elasticity characterizes the fraction of downstream momentum lost in a collision. If the streamwise velocity just prior to a collision is $u$, just after the collision it becomes $\varepsilon u$. The elasticity ranges from $\varepsilon=0$, representing a completely inelastic collision, to $\varepsilon=1$, representing a completely elastic collision.
\begin{figure}
	\centerline{\includegraphics{./figures/ch5/Fig1Concept.png}}
	\caption{Definition sketch of rarefied sediment transport with turbulent fluid drag and particle-bed collision forces. During saltation, pre-collisional streamwise velocities $u$ are transformed to postcollisional velocities $\ve u < u$.}
	\label{fig:fig1}
\end{figure}

Since the elasticity combines effects of particle shape and collision geometry and should vary from one collision to the next, the elasticity $\varepsilon$ is interpreted as a random variable, characterized by a statistical distribution $\rho(\varepsilon)$.
Some granular gas models have also included random elasticity \citep[e.g.][]{Serero2015}.

Assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep{Gordon1972}, the collision force in the downstream direction can be written as a Poisson pulse noise (sec. \ref{sec:einflux}):
\be F_C(u,t) = - m u \sum_{k=1}^{N_\nu(t)}(1-\varepsilon_k)\delta(t-\tau_k). \label{eq:col} \ee
Here, $N_\nu(t)$ is the number of collisions in time $t$ (a Poisson random variable), the $\tau_k$ ($k=1,2,\dots$) are times at which collisions occur, and the $\varepsilon_k$ are the elasticity coefficients, drawn from the distribution $\rho(\varepsilon)$ characterizing the fraction of momentum lost in each particle-bed collision.
This collision force is a sequence of random impulses which are proportional to the pre-collisional streamwise momentum. This collision model should be adequate when the contact times between moving and resting particles are small compared to the times between collisions. These conditions are always satisfied for the idealized saltation trajectories depicted in figure \ref{fig:fig1}.

\subsection{Turbulent fluid forces}
Fluid forces on a coarse particle in a viscous flow depend on the Reynolds number $\Re_p = d V/\nu$ defined by the particle size $d$, slip velocity $V$ between particle and fluid, and kinematic viscosity $\nu$.
These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Re_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
At realistic $\Re_p$, analytical results are inaccessible, so it is standard to include empirical corrections to the small $\Re_p$ formulas \citep{Schmeeckle2007,Clift1978}.
A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Re_p$ can be written $F_D = \frac{\pi}{8}
\rho_f d^2 C_D(\Re_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Re_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
In the present model I set $C_D = \frac{24}{\Rey_p}( 1 + 0.194 \Rey_p^{0.631})$ \citep{Clift1978,Gonzalez2017}, and I leave out acceleration and history terms for analytical tractability, although they are certainly relevant in coarse sediment transport \citep{Michaelides1997,Armenio2001}.

Drag forces have been argued to fluctuate rapidly compared to the inertial response times of submerged grains \citep{Fan2014}, and the magnitude of drag fluctuations has been observed to follow a Gaussian distribution \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
These ideas allow two major simplifications of the above drag force.
First, the drag can be split into quasi-steady and fluctuating components \citep{Michaelides1997}, and second, drag fluctuations can be modelled a Gaussian white noise, where the fluctuations are symmetric, the autocorrelation time is negligible, and the strength of fluctuations is described by a diffusivity $D$ \citep{Fan2014,Ancey2014}. 
Defining $\bar{V}$ as a representative slip velocity to be specified more carefully later,  $\bar{C}_D$ as the empirical drag coefficient evaluated at this slip velocity, and $\xi(t)$ as a Gaussian white noise of mean $0$ and variance $1$ \citep{Gardiner1983}, the fluid drag can be written
\be F_D(t) = \Gamma + \sqrt{2 D } \eta(t), \label{eq:drag}\ee
where $\Gamma = \frac{\pi}{8}
\rho_f d^2 \bar{C}_D \bar{V}^2$ is the steady component of the drag.

\subsection{Langevin equation for collisional bedload transport}

With the above forces, the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics becomes
\be m \dot{u}(t) = \Gamma + \sqrt{2D}\eta(t) - m u(t) \xi_{\nu, \ve}(t). \label{eq:langevin} \ee
This equation replaces the steady friction terms of earlier models with an episodic term, designed to provide a more realistic representation of particle-bed collisions.
Mathematically, equation \ref{eq:langevin} is a Langevin-like equation representing a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
Collisions introduce ``jumps" in velocity while turbulent generates "diffusion". The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.

Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Daly2010,Mau2014,Dubkov2019}.
One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
Here, the Ito interpretation (lower endpoint integration rule) is the physical choice the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional \citep[e.g.][]{Gardiner1983} since.
Given this integration rule, the remaining issues are to obtain the master equation characterizing the ensemble of velocities defined by \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.

@@

The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix section \ref{sec:langmasterderiv} with a limiting procedure, providing
\be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + I_c(u,t). \label{eq:master} \ee
In this equation, I introduced the scaled parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$ that scale the steady and fluctuating components of the fluid force against the collision rate $\nu$. The term
\be \mathcal{I}_c(u,t) = - P(u,t) + \int_0^1 \frac{d\ve}{\ve}P\big(\frac{u}{\ve},t \big) \rho(\ve) \label{eq:colint} \ee
is a ``collision integral" term representing particle-bed collisions.

Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bed load models \citep{Fan2014,Ancey2014}. 
Nonlocality is introduced by the collision integral \ref{eq:colint} which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
This term is analogous to the collision integral within the Boltzmann equation of kinetic theory \citep{Duderstadt1979, Brilliantov2004}. Physically, it corresponds to binary collisions between particles having different masses and random resitution coefficients \cite{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) diverges to infinity.
Mathematically, the collision integral represents the probability distribution of the product between the elasticity $\varepsilon$ and the downstream momentum $m u$ \citep[c.f.][]{Feller1968}.

Owing to its nonlocality, equation \ref{eq:master} does not admit analytical solutions as is, so further approximation is necessary.
Assuming the distribution of elasticity $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$, which is the case in experiments of rigid body collisions \citep{Glielmo2014}, allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
Expanding all terms in the integrand except $\rho(\varepsilon)$ provides
\be \mathcal{I}_c(u,t) = -P(u,t) + \frac{1}{\varepsilon'}P\big(\frac{u}{\ve'},t \big) + \sum_{k=1}^\infty \frac{\alpha_k }{k!}(\ve - \ve')^k \Big[\frac{1}{\ve}P\big(\frac{u}{\ve}\big)\Big]^{(k)}\Big|_{\ve=\ve'},\label{eq:expansion}\ee
where the $\alpha_k = \int_0^1 d\ve \rho(\ve) (\ve-\ve')^k $ are the central moments of $\ve$ around the mode elasticity $\ve'$ and the superscript $(k)$ denotes the $k$th derivative.

In what follows, I drop all but the first two terms in this expansion to obtain the leading order contribution of particle-bed collisions to the velocity distribution.
Higher orders could always be included later by perturbation theory \citep{Morse1953}.
I solve the resulting approximate equation in steady-state, when $\partial P(u,t)/\partial t = 0$. Equation \ref{eq:master} indicates that this solution will be a good approximation to the time-dependent problem when particle motions generally survive multiple collisions ($t\gg \nu^{-1}$).

\section{Results}
\label{sec:langresults}

\subsection{Derivation of the bedload velocity distribution}
\label{sec:langsolution}
Hereafter I drop the prime on the most common streamwise restitution coefficient $\ve'$.
With the truncation to two terms, equation \ref{eq:master} gives 
\be 0 = -\tilde{\Gamma}\partial_u P(u) + \tilde{D}\partial_u^2 P(u) -P(u) + \frac{1}{\ve} P\big(\frac{u}{\ve}\big),\label{eq:governer} \ee
which is an advanced functional differential equation. This equation is ``functional" in the sense that it is nonlocal in velocity due to the last term on the right hand side, and it is ``advanced" in that $u/\ve$ is advanced beyond the argument $u$ involved in the rest of this equation (since $0<\ve<1$).
Such equations seen some attention in the mathematics literature, where they are sometimes called pantograph equations \citep{Hall1989, Kim1998,Zaidi2015}.
